---
title: "Machine Learning: Course Project"
output: 
  html_document:
    keep_md: true
---  
### Author:  April Broome
### Date: 05-22-15  
## Overview:  
###  In this course project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell to predict the manner in which 6 participants performed barbell lifts. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. The training dataset contains the variable "classe" which is the outcome we are to predict with possible values: A, B, C, D, E.

## Read input files:
```{r CodeChunk1, echo=TRUE}
training <- read.csv("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",
                     na.strings=c("NA",""), stringsAsFactors = FALSE, header=TRUE)
testing <- read.csv("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",
                    na.strings=c("NA",""), stringsAsFactors = FALSE, header=TRUE)
```
## Clean the Data:  
### There are many columns in the training dataset which consist of mostly NA values.  To reduce processing time and improve the validity of the modeling, I determine which columns to keep by checking for more than 50% of the rows in that column populated (not NA).  I also removed columns 1-7 which contained data irrelevant to the analysis like X, subject name, timestamp, etc.
```{r CodeChunk2, echo=TRUE}
keeperCols <- which((colSums(!is.na(training)) >= 0.5*nrow(training)))
keeperCols <- keeperCols[8:60]
reducedTraining <- training[,keeperCols]
reducedTraining$classe <- as.factor(reducedTraining$classe)
reducedTesting <- testing[,keeperCols]
```

## Cross Validation:
### Initiate the Caret package and use the createDataPartition function to split out the original training dataset into a training set (75% of the data) and a testing (validation) set (25% of the data).  
```{r CodeChunk3, echo=TRUE, warning=FALSE, message=FALSE}
library(caret)
set.seed(100)
trainingSet <- createDataPartition(reducedTraining$classe, p=.75, list=F)
myTrain <- reducedTraining[trainingSet,]
myTest <- reducedTraining[-trainingSet,]
```

## Create a Machine Learning Model: 
### Random Forest Models have a reputation for being one of the best machine learning models so I use the "rf" method of the Caret package with the default settings.  Since creating the model takes such a long time, I check to see if an existing model file is available to use before running the "train" function.  
```{r CodeChunk4, echo=TRUE}
if (!file.exists("./rfFit.RData")) {
  rfFit <- train(classe ~., data = myTrain, method = "rf")
  save(rfFit, file="./rfFit.RData")
} else  {
  load("./rfFit.RData")
}
```

## Run a Prediction
### Using the Random Forest model I just created, I'll bring in the test (or validation) data reserved from the original training set to run against my machine learning model and generate a set of predicted values for "classe".
```{r CodeChunk6, echo=TRUE, warning=FALSE, message=FALSE}
myPrediction <- predict(rfFit, newdata=myTest)
```

## Out-of-Sample Error Analysis
### Based upon the reported accuracy of the Random Forest model, I estimate that my model will be at least 90% accurate.  To determine the actual accuracy of the machine learning model I created, I generate a Confusion Matrix to measure the error rate on the test (or out-of-sample) data.  The Confusion Matrix compares the estimates of classe that my machine learning model developed against the actual "classe"" values from the test data.
```{r CodeChunk7, echo=TRUE}
myConfMatrix <- confusionMatrix(myPrediction, myTest$classe)
myConfMatrix$overall[1]
```
### As you can see, the actual accuracy of my random forest model is 99%.

## Run model with Actual Test Data
### With a high accuracy rate from my model, I'm ready to run my model against the original test dataset provided.  I use the function provided to create individual files for each of 20 cases and use these files to submit for grading.
```{r CodeChunk8, echo=TRUE}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

answers <- predict(rfFit, newdata=reducedTesting)
pml_write_files(answers)
```